---
title: <strong>OCES4940A</strong>
subtitle: "No.1: Introduction to marine ecology and community dynamics"
author: "Masayuki Ushio<br>(ushio@ust.hk)"
date: "2023-07-11"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_section: true
toc-title: <br>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
```

<br>
<br>

# Basics
## Perform calculations by entering directly into the console

```{r}
1+1
10/2
5*5
print("Hello World")
```

## Assign values to variables
```{r}
a <- 10
b <- 5
a/b
```

## Data types
```{r}
#### Input data
num <- 10
chr <- "ABCDE"
vec <- c(1, 2, 3, 4, 5, 6)
lev <- factor(vec, levels=c(1,2,3,4,5,6))
mat <- matrix(vec, ncol=3)
lst <- list(mat1=mat, mat2=mat)
dat <- data.frame(mat)

# Show data type
class(num)
class(chr)
class(vec)
class(lev)
class(mat); class(lst); class(dat)

# Convert data type
num_chr <- as.character(num)
num_vec <- as.vector(num)
vec_mat <- as.matrix(vec)
dat_mat <- as.matrix(dat)
```

## Data generation
```{r}
data <- c(173, 167, 164, 174, 169, 174, 181, 162, 175, 155)
```


## Summary statistics
```{r}
summary(data)
mean(data)
median(data)
```


## Visualize data
```{r}
# Create a histogram
hist(data, main="Height Demo Data", xlab="Height (cm)", ylab="Frequency")
```

## Visualize data with various colors
```{r}
par(mfrow = c(2,2))
hist(data, main="Height Demo Data", col="grey")
hist(data, main="Height Demo Data", col="red3")
hist(data, main="Height Demo Data", col="green3")
hist(data, main="Height Demo Data", col="royalblue")
```


# Biodiversity
## Generate community data
```{r}
# Prepare each species abundance
rabit <- c(7, 10, 5, 8)
horse <- c(3, 0, 3, 1)
wolf <- c(0, 0, 2, 1)

# Generate data frame
df <- data.frame(sp1 = rabit, sp2 = horse, sp3 = wolf)
```


## Species richness
```{r}
richness <- rowSums(df > 0)
```

## Shannon index
$$
H'=-\sum_{i=1}^{R}{p_i \ln{p_i}}
$$
, where $p_i$ and $R$ indicate the relative abundance of species $i$ and species richness, respectively.


```{r}
# Calculate relative abundance
rel_df <- df/rowSums(df)

# Calculate Shannon index
## Community 1
### Simple summation
Sha_com1 <- - (( rel_df[1,1] * log(rel_df[1,1]) ) + ( rel_df[1,2] * log(rel_df[1,2]) ))

### For loop
Sha_com1 <- 0
for(i in 1:2) {
  Sha_com1 <- Sha_com1 - ( rel_df[1,i] * log(rel_df[1,i]) )
}

## Community 2
Sha_com2 <- - (rel_df[2,1] * log(rel_df[2,1]))

## Community 3
Sha_com3 <- 0
for(i in 1:3) {
  Sha_com3 <- Sha_com3 - (rel_df[3,i] * log(rel_df[3,i]))
}

## Community 4
Sha_com4 <- 0
for(i in 1:3) {
  Sha_com4 <- Sha_com4 - ( rel_df[4,i] * log(rel_df[4,i]) )
}

# Show all
Sha_com1; Sha_com2; Sha_com3; Sha_com4
```


## (Gini) Simpson index
$$
1-\lambda = 1-\sum_{i=1}^{R}{p_i^2}
$$

```{r}
## Community 1
simp_com1 <- 1 - sum(rel_df[1,]^2)
## Community 2
simp_com2 <- 1 - sum(rel_df[2,]^2)
## Community 3
simp_com3 <- 1 - sum(rel_df[3,]^2)
## Community 4
simp_com4 <- 1 - sum(rel_df[4,]^2)

# Show all
simp_com1; simp_com2; simp_com3; simp_com4
```


# Population dynamics
## Exponential Growth Model
$$
N_{t=0} = N_0　\\
N_{t+1} - N_t = rN_t
$$
```{r}
# Initial abundance
N0 <- 10

# Growth rate
r <- 0.5

# Set time limit
t_max <- 20

# Prepare an object that stores results
Nt <- rep(0, t_max)

# Growth model
Nt[1] <- N0
for (t in 1:t_max) {
  Nt[t+1] <- Nt[t] + r*Nt[t]
}

# Visualize the result
plot(Nt, type = "b")
```


## Discrete Logistic Model
$$
N_{t=0} = N_0　\\
N_{t+1} - N_t = rN_t(1-\frac{N_t}{K})
$$

```{r}
# Initial abundance
N0 <- 10

# Growth rate
r <- 0.5

# Carrying Capacity
K <- 100

# Set time limit
t_max <- 20

# Prepare an object that stores results
Nt <- rep(0, t_max)

# Growth model
Nt[1] <- N0
for (t in 1:t_max) {
  Nt[t+1] <- Nt[t] + r*Nt[t]*(1-Nt[t]/K)
}

# Visualize the result
plot(Nt, type = "b")
```


