---
title: <strong>OCES4940A</strong>
subtitle: "No.2: Fish population monitoring and its managemen<br><br> No.2-2: Data manipulation and plotting"
author: "<br>Masayuki Ushio<br>(ushio@ust.hk)"
date: "2023-07-13"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_section: true
toc-title: <br>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
```

<br>
<br>

# Reference
Contents in this tutorial generally follows "Introductory Fisheries Analyses with R" by Derek H. Ogle (2015). The data and R scripts are available at https://derekogle.com/IFAR/ and https://github.com/droglenc/IFAR (GPL-2.0 license)


# Preparations
## Load packages

```{r}
# Install a new package
#install.packages("FSA")
#install.packages("tidyverse")

# Load package
library(tidyverse) # tidyverse is a collection of useful packages
library(FSA)
```

## Set working directory
```{r, results='hide'}
# Check current working directory
getwd()

# Set working directory
# If you are using RStudio, your working directory will be your R file location
#setwd("XXXX")
```

## Load data
Please follow below:
1. Download demo data from https://derekogle.com/IFAR/scripts/inchBio.csv.
1. Create `data` folder in your working directly.
1. Place `inchBio.csv` in `data` folder.
1. Then, execute the following commands
```{r}
# Try some different functions
## read.table()
bio <- read.table("data/inchBio.csv", sep=",", header=TRUE)

## read.csv()
bio <- read.csv("data/InchBio.csv")

## readLines()
## (a bit unusual way...)
# tmp <- readLines("data/inchBio.csv",n=11)
# tmp2 <- tmp[1]
# for (i in 2:11) tmp2 <- rbind(tmp2,tmp[i])
# dimnames(tmp2) <- list(rep("", dim(tmp2)[1]), rep("", dim(tmp2)[2]))
# print(tmp2, quote=FALSE)
```


## Check data
```{r}
# Check structure
str(bio)

# Check head and tail of the data
headtail(bio,n=4)

# Load data in the FSA package
data(Mirex)

head(Mirex, n=3)
tail(Mirex, n=3)
```


# Data manipulations
## Data manipulation 1 using `tidyverse` 
```{r}
# filter() function: Select rows
bg <- filter(bio, species=="Bluegill")
head(bg, n = 3)

tmp <- filter(bio, species=="Yellow Perch", tl>=200)
head(tmp, n = 3)

## %in%
## "A %in% B" means "TRUE if A is in B".
tmp <- filter(bio, species %in% c("Black Crappie","Bluegill","Pumpkinseed"))

# select() function: Select columns
tmp <- select(bio, species, tl, w)
head(tmp, n = 3)

# rename() function: Change column names
tmp <- rename(bio,length = tl, weight = w)
head(tmp, n = 3)

# mutate() function: Add new columns
tmp <- mutate(bio, logL = log10(tl),logW = log10(w))
head(tmp, n = 3)

# arrange() function: Sort rows according to columns
tmp <- bio %>% arrange(tl) # Ascending order
head(tmp, n = 3)

tmp <- bio %>% arrange(desc(tl)) # Descending order
head(tmp, n = 3)

# %>% ("pipe" operator in magrittr package)
bg <- bio %>%
  filter(species == "Bluegill") %>%
  select(-tag, -scale) %>%
  mutate(logL = log10(tl),
         logW = log10(w),
         lcat = lencat(tl,w=10)) %>%
  arrange(netID,tl)

head(bg,n=3)

## lencat() creates "length category" and "w" is the width of the category
## Examples:
lencat(1:10, w = 1)
lencat(1:10, w = 2)
lencat(1:10, w = 5)

# Processing without %>%
## Need many temporal objects
tmp1 <- filter(bio, species == "Bluegill")
tmp2 <- select(tmp1, -tag, -scale)
tmp3 <- mutate(tmp2, logL = log10(tl), logW = log10(w), lcat = lencat(tl, w=10))
bg2 <- arrange(tmp3, netID, tl)

head(bg,n=3)
head(bg2,n=3)
```

## Data manipulation 2 using `tidyverse` 
### Load other data
1. Download demo data from https://derekogle.com/IFAR/scripts/inchGear.csv.
1. Download demo data from https://derekogle.com/IFAR/scripts/inchAge.csv.
1. Place `inchGear.csv` and `inchAge.csv` in `data` folder.
1. Then, execute the following commands
```{r}
# Load data
gear <- read.csv("data/inchGear.csv")

# Check data
head(gear,n=3)
head(bio,n=3)          # reminder of what bio looks like

# Manipulate data
bio2 <- left_join(gear, bio, by = "netID") %>%
  rename(effort = unitsOfEffort) %>%
  select(netID, netType, year, effort, fishID, species, tl)

headtail(bio2)


# Check what you did using left_join()
## Help
?left_join

## Individual process
left_join(gear, bio, by="netID") %>% headtail
left_join(gear, bio) %>% headtail
#left_join(gear %>% select(-netID), bio) %>% headtail


# Load and check data
age <- read.csv("data/inchAge.csv")
head(age,n=4)

# Merge data using merge() function
age2 <- merge(bio2, age) %>%
  select(-c(netID, netType, effort, ageCap))
head(age2,n=4)

# Check what you did using merge()
## Help
?merge

## Individual process
df1 <- data.frame(id = 1:10, a = 11:20)
df2 <- data.frame(id = 1:12, b = c(NA, 13:23))
merge(df1, df2)


# Other useful functions (will explain when it shows up)
#pivot_longer()
#pivot_wider()
#group_by()
#summarize()
```

**There are so many functions and pacakges, so it is impossible to remember and master everything. Just learn how to learn new functions and packages.**

# Plotting
## Install and load new pacakges
```{r}
#install.packages("FSAdata")
library(FSAdata)

#install.packages("plotrix")
library(plotrix)
```

## Load some internal data
```{r}
# Load some internal data
data(BullTroutRML1)
head(BullTroutRML1,n=3)

# The following data will not be used for plotting
#data(BullTroutRML2)
#head(BullTroutRML2,n=3)

#data(BloaterLH)
#head(BloaterLH,n=3)
```

## plot() function
```{r}
plot(mass ~ fl, data = BullTroutRML1)

# Different methods
#plot(BullTroutRML1$mass ~ BullTroutRML1$fl)
#plot(BullTroutRML1$fl, BullTroutRML1$mass)
```

```{r}
plot(mass ~ fl,
     data = BullTroutRML1,
     ylim = c(0, 1600), xlim = c(0, 500),
     ylab = "Weight (g)", xlab = "Fork Length (mm)")
```
```{r}
plot(mass ~ fl, # variables
     data = BullTroutRML1, # data
     ylim = c(0, 1600), xlim = c(0, 500), # plot range
     ylab = "Weight (g)", xlab = "Fork Length (mm)", # axis label
     pch = 19) # point type
```


## ggplot() function
```{r}
# Load ggplot2 library
#library(ggplot2) # Not necessary becuase it is included in tidyverse package

# Basic usage
ggplot(data = BullTroutRML1, aes(x = fl, y = mass)) +
  geom_point()

# Using pipe
BullTroutRML1 %>%
  ggplot(aes(x = fl, y = mass)) +
  geom_point()
```

```{r}
# Add layers
BullTroutRML1 %>%
  ggplot(aes(x = fl, y = mass)) +
  geom_point() +
  xlab("Fork length (mm)") +
  ylab("Weight (g)")
```
```{r}
# Add layers
BullTroutRML1 %>%
  ggplot(aes(x = fl, y = mass)) +
  geom_point() +
  stat_smooth() +
  xlab("Fork length (mm)") +
  ylab("Weight (g)")
```
