---
title: <strong>OCES4940A</strong>
subtitle: "No.2: Fish population monitoring and its managemen<br><br> No.2-1: R Basics, again"
author: "<br>Masayuki Ushio<br>(ushio@ust.hk)"
date: "2023-07-13"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_section: true
toc-title: <br>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
```

<br>
<br>

# Reference
Contents in this tutorial generally follows "Introductory Fisheries Analyses with R" by Derek H. Ogle (2015). The data and R scripts are available at https://derekogle.com/IFAR/ and https://github.com/droglenc/IFAR (GPL-2.0 license)


# R Basics, again
## Load FSA package (FSA: Simple Fisheries Stock Assessment Methods)

```{r}
# Install a new package
#install.packages("FSA")

# Load pacakge
library(FSA)
```

## Basic calculations
```{r}
# This is a whole-line comment
3+2*7  # This is a post-line comment

# Assign value to an object
tmp <- 2*6
tmp2 = 2*6 # Another expression

# Check the content
tmp
tmp2

# If you use a parenthesis, the command is executed and the result is displayed
( tmp <- 7+4*2 )

# Multiplication
tmp*3

# Logarithm
log10(1000)

log(1000)              # base= set by default (to exp(1))
log(1000,base=10)      # base= set explicitly, by name
log(1000,10)           # base= set explicitly, by position

# Display format
num <- 123456789
format(num,scientific=FALSE,digits=4)
format(num,scientific=TRUE,digits=4)
```

## Data types
```{r}
# Vector
( lens <- c(75,87,45,63,77,93) ) # Integer or numeric
( wts <- c(13,14.2,7.8,9,10.1,10.1) ) # Numeric
( gender <- c("M","F","F","M","M","F")) # Character
( mat <- c(TRUE,TRUE,FALSE,FALSE,FALSE,TRUE) ) # Logical

# Check class
class(lens)
class(wts)
class(gender)
class(mat)
```

## How to access each element
```{r}
# How to access each element
## Take the 3rd value
lens
lens[3]

## Take the 1st and 3rd values
lens[c(1,3)]

## Remove the 1st and 3rd values
lens[-c(1,3)]

## Other examples
hab <- c(35, 45, 20)
names(hab) <- c("rocky", "sandy", "silty")

## The 2nd element
hab[2]

## Specify the name of a value
hab["sandy"]
```


## Some calculations
```{r}
# Some calculations
lens/25.4
lens+10
lens^3

lens+wts
wts/(lens^3)*10000

lens + c(10,100)
lens + c(1,10,100,1000)
```

## Handling matrix and data frame
```{r}
# Manipulate data class 1
( gender1 <- factor(gender) )
class(gender1)

( gender2 <- factor(gender,levels=c("M","F")) )

as.numeric(gender1)
as.numeric(gender2)

# Manipulate data class 2
rbind(lens,wts)
( lw <- cbind(lens,wts) )

rbind(lens,gender)
rbind(lens,mat)
rbind(gender,mat)

# Access elements of a matrix
lw[[1,2]]          # 1st row, 2nd column (no name for element)
lw[c(1,3),2]       # 1st & 3rd rows, 2nd column
lw[-c(1,3),2]      # exclude 1st & 3rd rows, 2nd column

lw[1,]                            # 1st row
lw[,1]                            # 1st column
lw[,"wts"]                        # wts column

# Data frame
( df <- data.frame(lens,wts,gender,mat) )

## Access elements of a data frame
df[1,2]                           # 1st row, 2nd column
df[1,]                            # 1st row/individual
df[,2]                            # 2nd column/variable
df[,"wts"]                        # wts column/variable

df$lens
df$gender

class(df$lens)
df$lens[c(1,3)]
df$lens/25.4

log10(df$lens)
```

# Define a function, "for loop", and "if else" statements
## Define a function
```{r}
# Define a simple function
eqn1 <- function(x,y) {
  # simple calculation
  res <- x + y
  # return the result
  return(res)
}

eqn1(10, 100)   # 10 + 100
```


## For loop
```{r}
# Create vector of values
tmp <- c(7,2,3,5,4)

# Use loop to show the values in the vector
for(i in 1:5) {
  print(tmp[i])
}
```

## Logical statements
```{r}
z <- 1

# Equal to
z == 1

# Not equal to
z != 1

# Less than
z < 10

# Less than or equal to
z <= 10
z <= 1
z <= 0.99999
  
# Greater than
z > 10

# Greater than or equal to
z >= 10
z >= 1
z >= 0.99999
```



## If else statement
```{r}
# If statement
z <- 1


## method 1
if (z > 0) print("z is greater than 0")

## method 2
if (z > 0) { print("z is greater than 0") }

## method 3
if (z > 0) {
  print("z is greater than 0")
}


# If else statement
## method 1
if (z > 0) print("z is greater than 0") else print("z is not greater than 0")

# method 2
if (z > 0) { print("z is greater than 0") } else { print("z is not greater than 0") }

# method 3
if (z > 0) {
  print("z is greater than 0")
} else {
  print("z is not greater than 0")
  }


# If else if statement
if (z > 0) {
  print("z is greater than 0")
} else if (z == 0) {
  print("z is 0")
  } else {
  print("z is not greater than 0")
}
```